---
title: "Heart Disease Prediction Model"
author: "Tan Hong Yi 32061412"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    theme: flatly
    highlight: haddock
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Import Libraries

```{r echo=FALSE}
library(tidyverse)
library(dplyr)
library(visdat)
library(rpart)
library(rpart.plot)
```

## Dataset Overview

### Basic info

```{r}
# read dataset and define numerical columns and categorical columns
framingham <- read.csv("framingham.csv")
```

```{r}
# take a peek at the dataset
framingham
```

```{r}
# show columns
colnames(framingham)
```

### Summary for continuous data columns

```{r}
# define categorical and numerical columns
numerical_columns <- c('age', 'cigsPerDay', 'totChol', 'sysBP', 'diaBP', 'BMI', 'heartRate', 'glucose')
categorical_columns <- c('male', 'education', 'currentSmoker', 'BPMeds', 'prevalentStroke', 'prevalentHyp', 'diabetes')
label_column <- c('TenYearCHD')
```

```{r}
framingham[numerical_columns] %>% 
  summary()
```

### Summary for categorical data columns

'education' column has 4 categories(1,2,3,4) while other columns have 2 categories(0,1), and values in NA category indicates missing data.

```{r}
framingham[c(categorical_columns, label_column)] %>% 
  pivot_longer(everything()) %>%
  group_by(across(everything())) %>%
  summarise(N=n()) %>%
  pivot_wider(names_from = name,values_from=N)
```

## Exploratory data analysis

### Univariate Analysis

#### Continuous columns

```{r fig.height=8, fig.width=12}
framingham[numerical_columns] %>%
  gather(Variable, Value) %>%
  ggplot(aes(x=Value)) + 
  geom_histogram(aes(y=..density..), colour="black", fill="white")+
  geom_density(alpha=.2, fill="#FF6666") +
  geom_vline(aes(xintercept=mean(Value)), color="blue", linetype="dashed", size=2)+
  labs(title="Distribution of Continuous Variables",x="Value", y = "Density")+
  facet_wrap(~Variable, scales="free")
```

#### Categorical columns

```{r}
framingham[c(categorical_columns, label_column)] %>% 
  pivot_longer(everything()) %>%
  group_by(across(everything())) %>%
  summarise(N=n(), percentage=(n()/nrow(framingham))*100) %>%
  ggplot(aes(x=value, y=percentage, fill=value)) +
  geom_bar(stat="identity") +
  facet_wrap(~ name, ncol=3, scales="free") +
  labs(title="Distribution of Categorical Variables",x="Category", y = "Percentage(%)")+
  coord_flip()
```

## Data Pre-processing

### Missing value

In Basic Info under Dataset Overview section, missing data is noticed. Let's see whats missing and how many of them.

```{r}
framingham %>%
  vis_miss()+
  ggplot2::theme(legend.position = "bottom")
```

```{r}
colSums(is.na(framingham))
```

### Missing value treatment

```{r}
# remove rows with missing values
filtered_framingham <- framingham %>% 
  filter(if_all(everything(), Negate(is.na)))

# number of rows after filter
nrow(filtered_framingham)
```

### Model fitting

```{r}
filtered_framingham_fit <- rpart(TenYearCHD ~ ., data = filtered_framingham)

var_importance <- 
  tibble(variable = names(filtered_framingham_fit$variable.importance), importance = filtered_framingham_fit$variable.importance)
```

```{r fit-var-importance, echo=TRUE}
var_importance %>% 
  mutate(variable = fct_reorder(variable, importance)) %>% 
  ggplot(aes(x = importance, y = variable)) +
  geom_segment(aes(x = 0, y = variable, xend = importance, yend = variable)) +
  geom_point(colour = "red")
```

```{r}
filtered_framingham_fit_gini <- rpart(TenYearCHD ~ ., data = filtered_framingham, parms = list(split = "gini"))

printcp(filtered_framingham_fit_gini)
rpart.plot(filtered_framingham_fit_gini)
```

```{r}

pred_gini <- predict(filtered_framingham_fit_gini, filtered_framingham, type = "class")
table(filtered_framingham$TenYearCHD, pred_gini)
```

```{r echo=FALSE}
# install.packages("caTools")       # For sampling the dataset 
# install.packages("randomForest")
library(caTools) 
library(randomForest) 
filtered_framingham$TenYearCHD <- factor(filtered_framingham$TenYearCHD)
filtered_framingham <- filtered_framingham %>%
  select(TenYearCHD, everything())

split <- sample.split(filtered_framingham[c(categorical_columns, numerical_columns)], SplitRatio = 0.8) 
  
train <- subset(filtered_framingham, split == "TRUE") 
test <- subset(filtered_framingham, split == "FALSE") 

set.seed(120)  # Setting seed 

```

```{r}

classifier_RF = randomForest(x = train[-1],
                             y = train$TenYearCHD,
                             ntree = 500)
y_pred = predict(classifier_RF, newdata = test[-1]) 


plot(classifier_RF) 
  
# Importance plot 
importance(classifier_RF) 
  
# Variable importance plot 
varImpPlot(classifier_RF) 
```

```{r}
classifier_RF
```

```{r}
confusion_mtx = table(test[,1], y_pred) 
confusion_mtx 
```
